### 2018-04-11 paper/user-model.tex
rewrote HMM model

To this end, we use a Hidden Markov Model (HMM) of order one, with multiple discrete observation processes. 
The \emph{hidden} Markovian process $S=S_1,\dots,S_T$ represents the user's progression among the three proposed locations up to time step $T$.
For every device $d$ owned by the user, the \emph{observable} process $O_d=O_{d,1},\dots,O_{d,T}$ reads $d$'s state: it is either online ($o_d$) or offline ($\overline{o_d}$) at each point in time. 
The probability of observable events depends only on the current location.
%The \emph{observable} process $O=O_1,\dots,O_T$ reads ``toggle'' actions performed by the user on her devices. The probability of observable events depends on the current location.
For example, the probability that the user's home computer is on while she is at work is null (in our model, users never forget to turn off their fixed devices before leaving).

@@

$$A =
\kbordermatrix{
      & W            & O            & H            \cr
    W & \sfrac{2}{3} & \sfrac{1}{3} & 0            \cr
    O & \sfrac{1}{3} & \sfrac{1}{3} & \sfrac{1}{3} \cr
    H & 0            & \sfrac{1}{3} & \sfrac{2}{3} \\[0.3em]
}, \;
B = 
\kbordermatrix{
      & W    & O    & H    \cr
    t_n & 0.1  & 0.45 & 0.1  \cr
    t_p & 0.45 & 0.55 & 0.45 \cr
    t_w & 0.45 & 0    & 0    \cr
    t_h & 0    & 0    & 0.45 \\[0.3em]
}$$

@@

We consider a single user, owning a set of $\mathcal{D}$ devices.
Time is discretized here, and the user's activity is sequential: she can do zero or one action per ``round''of $T$s.
A HMM $\lambda_N$, comprising a set of states $\mathcal{S}$ and a set of observable events $\mathcal{O}$, is defined as follows:

@@

	\item $A = \left\{ a_{ij}\right\}_{(i,j)\in\mathcal{S}^2}$ is the transition probability matrix:\\
	$a_{ij}=P[S_t=j \mid S_{t-1}=i]$;

	\item $B = \left\{ b_{ik}\right\}_{(i,k)\in\mathcal{S}\times\mathcal{O}}$ is the matrix of states conditional densities:
	$b_{ik} = P[O_t=k \mid S_t = i]$.
\end{itemize}

In our case, the states are the different locations where the user is susceptible to go.
We considered the three following locations: $\mathcal{S}=\left\{ \mathit{work}, \mathit{outside}, \mathit{home} \right\}$. 
There is one action per device $d\in\mathcal{D}$, plus a ``none'' action when the user does not do anything: 
$\mathcal{O} = \left\{ t_d \right\}_{d\in \mathcal{D}} \bigcup \left\{ t_\text{none} \right\}$.
$t_d$ means that the user turned her device $d$ on or off (i.e. \emph{toggled} it), depending on the device's previous state.
\commentAL{TODO Rephrase:} We also note $o^{(d)}_t=1$ (resp. 0) when the user toggled (resp. did not toggle) device $d$ at time $t$.

In \cref{fig:hmm}, we show an example HMM of a user's behavioral HMM, along with its graphical representation (the initial state distribution $\Pi$ was left out for simplicity). 
The represented user possesses a phone, that she carries around with her, a home computer, that is only accessible from her home, and a workstation, located at her workplace.

@@

\caption{\label{fig:sample_usage}Sample device usage using the previous HMM specifications. 
%The bottom line reads the HMM's states: \textbf{W}ork, \textbf{O}utside, and \textbf{H}ome, while the dark lines read when the devices (\textbf{h}ome computer, \textbf{p}hone, and \textbf{w}orkstation) are online.
}

@@

Using the toggle events, we compute the device's \emph{connection} events, represented by the multivariate Bernoulli process $C=\left\{ c^{(d)} \right\}_{d\in\mathcal{D}}$. 
$c^{(d)}_t=1$ (resp. 0) means that the device $d$ was online (resp. offline) at time step $t$.

We state that every device starts offline, and compute $c^{(d)}_t$ as follows:

$$ \forall d \in \mathcal{D},
\begin{cases}
c^{(d)}_t=0 &\text{when } t=0; \\
c^{(d)}_t=c^{(d)}_{t-1} \oplus o^{(d)}_t &\text{when } t>0.
\end{cases}$$

Figure~\ref{fig:sample_usage}, shows a possible trace of the user's behavior, using the parameters depicted in figure~\ref{fig:hmm}. It read the timeline of $c^{(d)}$ for each device $d$. On the bottom of the graph, the first letter of the user's location is shown at each time step. We see that any number of devices can be connected at the same time.

@@

\commentAL{Need to explain how devices go from sequence to the HMM at some point. 
We first define the prediction of behavior from the HMM directly.}

\subsubsection{Predicting future behavior} % (fold)
\label{ssub:predicting_future_behavior}
\commentAL{Why we need to make these predictions is not explained yet.}
Devices want to predict whether they are likely to be connected in the near future.
This information will be publicly available, and allow routes prediction for the upcoming files exchange.

Given the model $\lambda_N$, the current user's location $S_t$ and the current connection state of the devices $C_t$, we can predict the devices connections in $k$ round $C_{t+k}$.

If the device $d$ is connected at time $t$ (resp. disconnected), the probability that $d$ is connected at time $t+k$ is equal to the probability that the user toggles $d$ an even (resp. odd) number of times between round $t+1$ and $t+k$:

$$
P\left[c^{(d)}_{t+k}=1 \mid c^{(d)}_{t}=1 ,S_t \right]= P\left[ \sum\limits_{i\in[\![ 1,k]\!]}o^{(d)}_{t+i} \text{ is even} \mid S_t\right] 
$$

To compute this probability, one thus need to predict the probability that the user toggles $d$ at each round until $t+k$. This is done recursively:

\begin{multline*}
P\left[o^{(d)}_{t+i}=1 \mid S_{t} \right] = \\
\sum\limits_{s \in \mathcal{S}} P\left[ S_{t+1}=s \mid S_t\right] * P\left[o^{(d)}_{t+i}=1 \mid S_{t+1}=s \right]
\end{multline*}

Knowing that the probability of toggling $d$ during the next round is computed as follows:

\begin{multline*}
P\left[o^{(d)}_{t+1}= 1 \mid S_{t} \right]=\\
\sum\limits_{s \in \mathcal{S}} P\left[ S_{t+1}=s \mid S_{t}\right] * P\left[ o^{(d)}_{t+1}=1 \mid S_{t+1}=s\right] 
\end{multline*}

Finally, we enumerate the set $\mathcal{B}^{(k)}_\text{even}$ (resp. $\mathcal{B}^{(k)}_\text{even}$) of sequences of $k$ booleans having an even (resp. odd) number of true values, and compute the probability that device $d$ will be connected in $k$ turns, knowing that it is connected (resp. disconnected) at time $t$:

\begin{multline*}
P\left[c^{(d)}_{t+k}=1 \mid c^{(d)}_{t}=1 ,S_t \right]=\\
\sum\limits_{B \in \mathcal{B}_\text{even}}\sum\limits_{i\in[\![ 1,k]\!]} P\left[o^{(d)}_{t+i}= B\left[i\right] \mid S_{t} \right]
\end{multline*}

\begin{multline*}
P\left[c^{(d)}_{t+k}=1 \mid c^{(d)}_{t}=0 ,S_t \right]=\\
\sum\limits_{B \in \mathcal{B}_\text{odd}}\sum\limits_{i\in[\![ 1,k]\!]} P\left[o^{(d)}_{t+i}= B\left[i\right] \mid S_{t} \right]
\end{multline*}

% \begin{multline}
% P\left[o^{(d)}_{t+1}= 1 \mid S_{t} \right]= P\left[o^{(d)}_{t+i+1}= 1 \mid S_{t+i} \right] = \\
% \sum\limits_{s \in \mathcal{S}} P\left[ S_{t+1}=s \mid S_{t}\right] * P\left[ o^{(d)}_{t+1}=1 \mid S_{t+1}=s\right] \\

\Cref{fig:sample_usage} plots a sample device usage timeline using the previous model. 
Each device starts offline; the user can toggle one device per round (or do nothing).
In this figure, we show the device \emph{usage}, that is a consequence of the $\mathit{toggle}$ events defined by the HMM.
As a consequence, several devices can be online at the same time, as happens in the figure.

% Using the matrices notations:

% $$
% \begin{cases}
% P\left[o^{(d)}_{t+i+1}=1 \mid S_{t+i} = i\right] = \sum\limits_{j \in \mathcal{S}} a_{ij} * b_{jk}\\

% P\left[o^{(d)}_{t+i}=1 \mid S_{t} = i \right] = \sum\limits_{j \in \mathcal{S}} a_{ij} * P\left[o^{(d)}_{t+i}=1 \mid S_{t+1}=s \right]
% \end{cases}
% $$

% subsubsection predicting_future_behavior (end)

>>>>

To this end, we use a Hidden Markov Model (HMM) of order one, with multiple discrete observation processes. 
The hidden Markovian process $S$ represents the user's locations, while the \emph{observable} process $O=O_1, \dots, \O_T$ represents the connection state of her devices (either connected or disconnected at any point in time).
% The \emph{hidden} Markovian process $S=S_1,\dots,S_T$ represents the user's progression among the three proposed locations up to time step $T$.
% For every device $d$ owned by the user, the \emph{observable} process $O_d=O_{d,1},\dots,O_{d,T}$ reads $d$'s state: it is either online ($o_d$) or offline ($\overline{o_d}$) at each point in time. $O_t = O_1, \dots, \O_T$ represents the connection of all the user's devices, such that $o_d \in O_t$ means that device $d$ was connected at time $t$.

% The probability of observable events depends only on the current location.
% For example, the probability that the user's home computer is on while she is at work is null (in our model, users never forget to turn off their fixed devices before leaving).

@@

$$A =
\kbordermatrix{
      & W            & O            & H            \cr
    W & \sfrac{2}{3} & \sfrac{1}{3} & 0            \cr
    O & \sfrac{1}{3} & \sfrac{1}{3} & \sfrac{1}{3} \cr
    H & 0            & \sfrac{1}{3} & \sfrac{2}{3} \\[0.3em]
}, \;
B = 
\kbordermatrix{
      & W     & O   & H   \cr
    o_p & 0.7 & 0.6 & 0.8 \cr
    o_w & 0.7 & 0   & 0   \cr
    o_h & 0   & 0   & 0.7 \cr
    o_l & 0.2 & 0.4 & 0.6 \\[0.3em]
}$$

@@

We consider a single user, owning a set of $\mathcal{D}$ devices.
Time is discretized, and the devices' observations are independent: Alice can have her laptop and phone connected at the same time.
The hidden process $S=S_1,\dots,S_T$ is built from an alphabet of states $\mathcal{S}$, while the observable process $O=O_1, \dots, \O_T$ contains events from the alphabet $\mathcal{O}$. We define a multivariate HMM $\lambda_N$ as follows:

@@

	\item $A = \left\{ a_{i\,j}\right\}_{(i,j)\in\mathcal{S}^2}$ is the state transition matrix:\\
	$a_{i\,j}=P[S_t=j \mid S_{t-1}=i]$;

	\item $B = \left\{ b_{i\,k}\right\}_{(i,k)\in\mathcal{S}\times\mathcal{O}}$ is the matrix of event emission probabilities for each state:
	$b_{i\,k} = P[ k \in O_t \mid S_t = i]$.
\end{itemize}

In our case, the states are the different locations where the user is susceptible to go;
we considered the three following ones: $\mathcal{S}=\left\{ \mathit{work}, \mathit{outside}, \mathit{home} \right\}$. 
There is one event per device $d$: $\mathcal{O} = \left\{ o_d \right\}_{d\in \mathcal{D}}$. Several devices can be connected at the same time, such that $O_t$ is a set, and $o_d \in O_t$ means that device $d$ was connected at time $t$. Devices' connection probabilities are independent.

In \cref{fig:hmm}, we show an example HMM of a user's behavioral HMM, along with its graphical representation (the initial state distribution $\Pi$ was left out for simplicity). 
The represented user possesses a phone and a laptop, that she carries around with her, a home computer, that is only accessible from her home, and a workstation, located at her workplace. She never forgets to turn off her fixed appliances before leaving.

@@

\caption{\label{fig:sample_usage}Sample device usages using the previous HMM specifications. 
%The bottom line reads the HMM's states: \textbf{W}ork, \textbf{O}utside, and \textbf{H}ome, while the dark lines read when the devices (\textbf{h}ome computer, \textbf{p}hone, and \textbf{w}orkstation) are online.
}

@@

% Using the toggle events, we compute the device's \emph{connection} events, represented by the multivariate Bernoulli process $C=\left\{ c^{(d)} \right\}_{d\in\mathcal{D}}$. 
% $c^{(d)}_t=1$ (resp. 0) means that the device $d$ was online (resp. offline) at time step $t$.

% We state that every device starts offline, and compute $c^{(d)}_t$ as follows:

% $$ \forall d \in \mathcal{D},
% \begin{cases}
% c^{(d)}_t=0 &\text{when } t=0; \\
% c^{(d)}_t=c^{(d)}_{t-1} \oplus o^{(d)}_t &\text{when } t>0.
% \end{cases}$$

Figure~\ref{fig:sample_usage}, shows a possible trace of the user's behavior, using the parameters depicted in figure~\ref{fig:hmm}. 
It reads the timeline of $o_{d}$ for each device $d$. On the bottom of the graph, the first letter of the user's location is shown at each time step.
We see that devices connections depend on the user's location, and that any number of devices can be connected at the same time.

@@

\subsubsection{Predicting future behavior} % (fold)
\label{ssub:predicting_future_behavior}
\commentAL{Why we need to make these predictions is not explained yet.}
Devices want to predict whether they are likely to be connected in the near future.
This information will be publicly available, and will allow routes selection for the  files exchange between users.

Because devices connections only depend on the users location, we can predict the devices connections in $i$ round $O_{t+i}$ given only the model $\lambda_N$ and the current user's location $S_t$. We first compute the probability that a device $d$ will be connected next turn:

$$
P\left[ o_d \in O_{t} \mid S_{t-1} = s \right] = 
\sum\limits_{s' \in \mathcal{S}} 
a_{s\,s'} * b_{s'\,o_d}.
$$

It is the sum of the probabilities, for each state $s$, that the user switches to $s$ and uses her device.
Recursively, predicting $i$ turns in advance follows the same logic:

\begin{multline*}
P\left[ o_d \in O_{t+i} \mid S_{t} = s \right] = \\
\sum\limits_{s' \in \mathcal{S}}
P\left[ S_{t+1} = s' \mid S_t = s \right] * 
P\left[ o_d \in O_{t+i} \mid S_{t+1} = s'\right].
\end{multline*}
